---
title: "bioLeak: Leakage-Aware Biomedical Modeling"
author: "Selçuk Korkmaz"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{bioLeak: Leakage-Aware Biomedical Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE
)
```

## Why bioLeak

bioLeak is a leakage-aware modeling toolkit for biomedical and machine-learning
analyses. Its purpose is to prevent and diagnose information leakage across
resampling workflows where training and evaluation data are not truly
independent because samples share subjects, batches, studies, or time.

Standard workflows are often insufficient. Random, row-wise cross-validation
assumes samples are independent. Global preprocessing (imputation, scaling,
feature selection) done before resampling lets test-fold information shape the
training process. These choices inflate performance and can lead to incorrect
biomarker discovery, misleading clinical signals, or models that fail in
deployment.

Data leakage means any direct or indirect use of evaluation data in training or
feature engineering. In biomedical data, leakage commonly appears as:
- repeated measurements from the same patient split across folds
- batch or study effects that align with the outcome
- duplicates or near-duplicates across train and test
- time-series lookahead or random splits that use future information
- preprocessing that uses all samples instead of training-only statistics

bioLeak addresses these issues with leakage-aware splitting, guarded
preprocessing that is fit only on training data, and audit diagnostics that
surface overlaps, confounding, and duplicates.

## Guided workflow

The sections below walk through a leakage-aware workflow from data setup to
audits. Each step includes a leaky failure mode and a corrected alternative.

### Example data

```{r example-data}
library(bioLeak)

set.seed(123)
n <- 160
subject <- rep(seq_len(40), each = 4)
batch <- sample(paste0("B", 1:6), n, replace = TRUE)
study <- sample(paste0("S", 1:4), n, replace = TRUE)
time <- seq_len(n)

x1 <- rnorm(n)
x2 <- rnorm(n)
x3 <- rnorm(n)
linpred <- 0.7 * x1 - 0.4 * x2 + 0.2 * x3 + rnorm(n, sd = 0.5)
p <- stats::plogis(linpred)
outcome <- factor(ifelse(runif(n) < p, "case", "control"),
                  levels = c("control", "case"))

df <- data.frame(
  subject = subject,
  batch = batch,
  study = study,
  time = time,
  outcome = outcome,
  x1 = x1,
  x2 = x2,
  x3 = x3
)

df_leaky <- within(df, {
  leak_subject <- ave(as.numeric(outcome == "case"), subject, FUN = mean)
  leak_batch <- ave(as.numeric(outcome == "case"), batch, FUN = mean)
  leak_global <- mean(as.numeric(outcome == "case"))
})

df_time <- df
df_time$leak_future <- c(as.numeric(df_time$outcome == "case")[-1], 0)
predictors <- c("x1", "x2", "x3")


# Example data (first 6 rows)
head(df)

# Outcome class counts
as.data.frame(table(df$outcome))
```

The table preview shows the metadata columns (`subject`, `batch`, `study`, `time`),
the binary `outcome`, and three numeric predictors. The class count table
is the baseline prevalence for the `case` vs `control` labels; this is the
reference distribution we try to preserve with stratified splits.

### Create leakage-aware splits with `make_splits()`

`make_splits()` is the foundation. It returns a `LeakSplits` object with
explicit train/test indices and metadata. It assumes that the grouping columns
you provide are complete and that samples sharing a group must not cross folds.
Misuse to avoid:
- leaving `group = NULL` when subjects repeat
- using the wrong grouping column (for example batch instead of subject)
- using time-series CV with unsorted or missing time values
- relying on stratification when the outcome is missing or has only one class

**Leaky example: row-wise CV when subjects repeat**

```{r leaky-splits}
leaky_splits <- make_splits(
  df,
  outcome = "outcome",
  mode = "subject_grouped",
  group = NULL,
  v = 5,
  repeats = 1,
  stratify = TRUE
)

cat("Leaky splits summary (sample-wise CV):\n")
leaky_splits
```

The printed `LeakSplits` summary reports the split mode, number of folds, and
fold-level train/test sizes. Because `group = NULL`, each sample is treated as
its own group, so repeated subjects can cross folds and create leakage.

**Leakage-safe alternative: group by subject**

```{r safe-splits}
safe_splits <- make_splits(
  df,
  outcome = "outcome",
  mode = "subject_grouped",
  group = "subject",
  v = 5,
  repeats = 1,
  stratify = TRUE,
  seed = 10
)

cat("Leakage-safe splits summary (subject-grouped CV):\n")
safe_splits
```

Here, each subject is confined to a single fold. The fold counts should be
similar in size because `stratify = TRUE` balances the outcome across folds.
Use these fold sizes when reporting how many subjects/samples were held out.

**Other leakage-aware modes**

```{r splits-other-modes}
batch_splits <- make_splits(
  df,
  outcome = "outcome",
  mode = "batch_blocked",
  batch = "batch",
  v = 4,
  stratify = TRUE
)

cat("Batch-blocked splits summary:\n")
batch_splits

study_splits <- make_splits(
  df,
  outcome = "outcome",
  mode = "study_loocv",
  study = "study"
)

cat("Study leave-one-out splits summary:\n")
study_splits

time_splits <- make_splits(
  df,
  outcome = "outcome",
  mode = "time_series",
  time = "time",
  v = 4,
  horizon = 2
)

cat("Time-series splits summary:\n")
time_splits

nested_splits <- make_splits(
  df,
  outcome = "outcome",
  mode = "subject_grouped",
  group = "subject",
  v = 3,
  nested = TRUE
)

cat("Nested CV splits summary:\n")
nested_splits
```

Each summary shows the number of folds and fold sizes for that split strategy.
Use the batch and study summaries to verify that entire batches/studies are
held out, and use the time-series summary to check that early folds have
smaller training sizes due to the horizon constraint.

Assumptions and intent by mode:
- `subject_grouped`: keeps all samples from a subject together
- `batch_blocked`: keeps entire batches or centers together
- `study_loocv`: holds out each study in turn for external validation
- `time_series`: rolling-origin splits with optional `horizon` to prevent lookahead

Use `repeats` for repeated CV, `stratify = TRUE` to balance outcomes, and
`nested = TRUE` to attach inner folds. For large datasets, `progress = TRUE`
reports progress, and explicit indices can be memory intensive.

#### Scalability

**Handling Large Datasets (Compact Mode)**

For large datasets (e.g., $N > 50,000$) with many repeats, storing explicit
integer indices for every fold can consume gigabytes of memory. Use `compact = TRUE`
to store a lightweight vector of fold assignments instead. `fit_resample()` will
automatically reconstruct the indices on the fly.

```{r compact-splits}
# Efficient storage for large N
big_splits <- make_splits(
  df,
  outcome = "outcome",
  mode = "subject_grouped",
  group = "subject",
  v = 5,
  compact = TRUE  # <--- Saves memory
)

cat("Compact-mode splits summary:\n")
big_splits
cat(sprintf("Compact storage enabled: %s\n", big_splits@info$compact))
```

The summary is identical to a regular split, but the underlying storage is a
compact fold-assignment vector. Use the `compact` flag to confirm the memory-
saving mode is active.

### Guarded preprocessing and imputation

`bioLeak` uses guarded preprocessing to prevent leakage from global imputation,
scaling, and feature selection. The low-level API is:
- `.guard_fit()` to fit preprocessing on training data only
- `predict_guard()` to apply the trained guard to new data
- `.guard_ensure_levels()` to align factor levels across train/test
- `impute_guarded()` as a convenience wrapper for leakage-safe imputation

`.guard_fit()` fits a pipeline that can winsorize, impute, normalize, filter,
and select features. Assumptions and misuse to avoid:
- `fs = "ttest"` requires a binary outcome with enough samples per class
- `fs = "lasso"` requires `glmnet`
- `impute = "mice"` is not supported for guarded prediction
- `impute = "none"` with missing values will trigger a median fallback and
  missingness indicators to avoid errors

Supported preprocessing options include imputation (`median`, `knn`,
`missForest`, `none`), normalization (`zscore`, `robust`, `none`), filtering by
variance or IQR, and feature selection (`ttest`, `lasso`, `pca`). In
`impute_guarded()`, `winsor` and `winsor_thresh` control MAD-based clipping.

**Leaky example: global scaling before CV**

```{r leaky-scaling}
df_leaky_scaled <- df
df_leaky_scaled[predictors] <- scale(df_leaky_scaled[predictors])
scaled_summary <- data.frame(
  feature = predictors,
  mean = colMeans(df_leaky_scaled[predictors]),
  sd = apply(df_leaky_scaled[predictors], 2, stats::sd)
)
scaled_summary$mean <- round(scaled_summary$mean, 3)
scaled_summary$sd <- round(scaled_summary$sd, 3)

# Leaky global scaling: means ~0 and SDs ~1 computed on all samples
scaled_summary
```

The summary shows that scaling used the full dataset, so test-fold statistics
influenced the training transformation. This violates the train/test separation
and can bias performance estimates.

**Leakage-safe alternative: fit preprocessing on training only**

```{r guarded-preprocess}
fold1 <- safe_splits@indices[[1]]
train_x <- df[fold1$train, predictors]
test_x <- df[fold1$test, predictors]

guard <- .guard_fit(
  X = train_x,
  y = df$outcome[fold1$train],
  steps = list(
    impute = list(method = "median", winsor = TRUE),
    normalize = list(method = "zscore"),
    filter = list(var_thresh = 0, iqr_thresh = 0),
    fs = list(method = "none")
  ),
  task = "binomial"
)

train_x_guarded <- predict_guard(guard, train_x)
test_x_guarded <- predict_guard(guard, test_x)

cat("GuardFit object:\n")
guard
cat("\nGuardFit summary:\n")
summary(guard)

# Guarded training data (first 6 rows)
head(train_x_guarded)

# Guarded test data (first 6 rows)
head(test_x_guarded)
```

The `GuardFit` object records the preprocessing steps and the number of
features retained after filtering. The summary reports how many features were
removed and the preprocessing audit trail. The guarded train/test previews show
that missing values were imputed and (if requested) scaled using training-only
statistics; the test data never influences these values.

**Factor level guard (advanced)**

```{r guard-ensure-levels}
raw_levels <- data.frame(
  site = c("A", "B", "B"),
  status = c("yes", "no", "yes"),
  stringsAsFactors = FALSE
)

level_state <- .guard_ensure_levels(raw_levels)

# Aligned factor data with consistent levels
level_state$data

# Levels map
level_state$levels
```

The returned `data` keeps factor levels consistent across folds, while the
`levels` list records the training-time levels (including any dummy levels added
to preserve one-hot columns). Use these when transforming new data to avoid
misaligned model matrices.

**Leaky example: imputation using train and test together**

```{r leaky-impute}
train <- data.frame(a = c(1, 2, NA, 4), b = c(NA, 1, 1, 0))
test <- data.frame(a = c(NA, 5), b = c(1, NA))

all_median <- vapply(rbind(train, test),
                     function(col) median(col, na.rm = TRUE),
                     numeric(1))
train_leaky <- as.data.frame(Map(function(col, m) { col[is.na(col)] <- m; col },
                                 train, all_median))
test_leaky <- as.data.frame(Map(function(col, m) { col[is.na(col)] <- m; col },
                                test, all_median))

# Leaky medians computed on train + test
data.frame(feature = names(all_median), median = all_median)

# Leaky-imputed training data
train_leaky

# Leaky-imputed test data
test_leaky
```

The medians above were computed using both train and test data, so the test set
influences the imputation values. This is a classic leakage pathway because
test information directly alters the training features.

**Leakage-safe alternative: `impute_guarded()`**

```{r safe-impute}
imp <- impute_guarded(
  train = train,
  test = test,
  method = "median",
  winsor = FALSE
)

# Guarded-imputed training data
imp$train

# Guarded-imputed test data
imp$test
```

Here the imputation statistics are computed from the training set only. The
test set is transformed using those fixed values, preserving a clean separation
between training and evaluation data.

### Fit and resample with `fit_resample()`

`fit_resample()` combines leakage-aware splits with guarded preprocessing and
model fitting. It supports:
- built-in learners (`glmnet`, `ranger`)
- `parsnip` model specifications (recommended)
- custom learners via `custom_learners` (advanced)
- multiple metrics, class weights, and a positive class override

Assumptions and misuse to avoid:
- outcome must be binary (classification) or numeric (regression)
- `positive_class` must be a single value that exists in the outcome levels
- if you supply a matrix without metadata, you must remove leakage columns
  yourself (group, batch, study, time)

Use `learner_args` to pass model-specific arguments. For custom learners
(advanced), you can supply separate `fit` and `predict` argument lists. Set
`parallel = TRUE` to use future.apply for fold-level parallelism when available.
When `learner` is a
parsnip specification, `learner_args` and `custom_learners` are ignored.

Metrics used in this vignette:
- **AUC**: area under the ROC curve (0.5 = random, 1.0 = perfect; higher is better)
- **PR AUC**: area under the precision-recall curve (0 to 1; higher is better)
- **Accuracy**: fraction of correct predictions (0 to 1; higher is better)
- **RMSE**: root mean squared error (0 to infinity; lower is better)

Always report the mean and variability across folds, not a single fold value.

**Parsnip model specification (recommended)**

```{r parsnip-spec}
spec <- parsnip::logistic_reg() |>
  parsnip::set_engine("glm")
```

This spec uses base R `glm` under the hood and does not require extra model
packages. Use it in the examples below; custom learners are covered in the
Advanced section.

**Leaky example: leaky features and row-wise splits**

```{r fit-leaky}
fit_leaky <- fit_resample(
  df_leaky,
  outcome = "outcome",
  splits = leaky_splits,
  learner = spec,
  metrics = c("auc", "accuracy"),
  preprocess = list(
    impute = list(method = "median"),
    normalize = list(method = "zscore"),
    filter = list(var_thresh = 0),
    fs = list(method = "none")
  )
)

cat("Leaky fit summary:\n")
summary(fit_leaky)
metrics_leaky <- as.data.frame(fit_leaky@metric_summary)
num_cols <- vapply(metrics_leaky, is.numeric, logical(1))
metrics_leaky[num_cols] <- lapply(metrics_leaky[num_cols], round, digits = 3)

# Leaky fit: mean and SD of metrics across folds
metrics_leaky
```

The summary reports cross-validated performance. AUC ranges from 0.5 (random)
to 1.0 (perfect), while accuracy is the fraction of correct predictions.
Because the splits are leaky, these metrics can be artificially high and should
not be used for reporting.

**Leakage-safe alternative: grouped splits and clean predictors**

```{r fit-safe}
fit_safe <- fit_resample(
  df,
  outcome = "outcome",
  splits = safe_splits,
  learner = spec,
  metrics = c("auc", "accuracy"),
  preprocess = list(
    impute = list(method = "median"),
    normalize = list(method = "zscore"),
    filter = list(var_thresh = 0),
    fs = list(method = "none")
  ),
  positive_class = "case",
  class_weights = c(control = 1, case = 1),
  refit = TRUE
)

cat("Leakage-safe fit summary:\n")
summary(fit_safe)
metrics_safe <- as.data.frame(fit_safe@metric_summary)
num_cols <- vapply(metrics_safe, is.numeric, logical(1))
metrics_safe[num_cols] <- lapply(metrics_safe[num_cols], round, digits = 3)

# Leakage-safe fit: mean and SD of metrics across folds
metrics_safe

# Per-fold metrics (first 6 rows)
head(fit_safe@metrics)
```

`fit_resample()` returns a `LeakFit` object. Use `summary(fit_safe)` for a
compact report and inspect `fit_safe@metrics` and `fit_safe@predictions` for
details. When `refit = TRUE`, the final model and preprocessing state are stored
in `fit_safe@info$final_model` and `fit_safe@info$final_preprocess`.

The mean +/- SD table is the primary performance summary to report, while the
per-fold metrics reveal variability and potential instability across folds.

The examples above use a parsnip model specification. To swap in other models,
replace `spec` with another parsnip spec (see the gradient boosting example
below).

**Passing learner-specific arguments (optional)**

```{r learner-args}
if (requireNamespace("glmnet", quietly = TRUE)) {
  fit_glmnet <- fit_resample(
    df,
    outcome = "outcome",
    splits = safe_splits,
    learner = "glmnet",
    metrics = "auc",
    learner_args = list(glmnet = list(alpha = 0.5)),
    preprocess = list(
      impute = list(method = "median"),
      normalize = list(method = "zscore"),
      filter = list(var_thresh = 0),
      fs = list(method = "none")
    )
  )
  cat("GLMNET summary with learner-specific arguments:\n")
  summary(fit_glmnet)
} else {
  cat("glmnet not installed; skipping learner_args example.\n")
}
```

This summary reflects the same guarded preprocessing but a different model
configuration (here, `alpha = 0.5`). Use the mean +/- SD metrics to compare
learner settings under identical splits.

**SummarizedExperiment inputs (optional)**

```{r se-example}
if (requireNamespace("SummarizedExperiment", quietly = TRUE)) {
  se <- SummarizedExperiment::SummarizedExperiment(
    assays = list(counts = t(as.matrix(df[, predictors]))),
    colData = df[, c("subject", "batch", "study", "time", "outcome"), drop = FALSE]
  )

  se_splits <- make_splits(
    se,
    outcome = "outcome",
    mode = "subject_grouped",
    group = "subject",
    v = 3
  )

  se_fit <- fit_resample(
    se,
    outcome = "outcome",
    splits = se_splits,
    learner = spec,
    metrics = "auc"
  )
  cat("SummarizedExperiment fit summary:\n")
  summary(se_fit)
} else {
  cat("SummarizedExperiment not installed; skipping SE example.\n")
}
```

The summary is identical in structure to the data.frame case because
`fit_resample()` extracts predictors and metadata from the `SummarizedExperiment`.

### Advanced: Using Gradient Boosting with Parsnip

`bioLeak` natively supports `tidymodels` specifications. You can pass a `parsnip`
model specification directly to `fit_resample()`. This allows you to use
state-of-the-art algorithms like XGBoost, LightGBM, or SVMs while ensuring all
preprocessing remains guarded.

```{r fit-xgboost}
if (requireNamespace("parsnip", quietly = TRUE) &&
    requireNamespace("xgboost", quietly = TRUE)) {

  # 1. Define the model spec (no wrappers needed!)
  xgb_spec <- parsnip::boost_tree(
    mode = "classification",
    trees = 500,
    tree_depth = 6,
    learn_rate = 0.01
  ) |>
    parsnip::set_engine("xgboost")

  # 2. Fit with guards
  # Note: bioLeak automatically handles the training/test splits and guards
  fit_xgb <- fit_resample(
    df,
    outcome = "outcome",
    splits = safe_splits,
    learner = xgb_spec,   # Pass the spec directly
    metrics = "auc",
    preprocess = list(
      impute = list(method = "median"),
      # XGBoost handles interactions, but we still guard global stats
      normalize = list(method = "none"),
      filter = list(var_thresh = 0)
    )
  )

  cat("XGBoost parsnip fit summary:\n")
  summary(fit_xgb)
} else {
  cat("parsnip or xgboost not installed; skipping gradient boosting example.\n")
}
```

The summary reports cross-validated AUC for a non-linear gradient boosting
model. Use the mean +/- SD table for reporting, and compare it to linear models
to assess whether non-linear signal improves predictive performance without
introducing leakage.

### Advanced: Custom learners

Use custom learners when a model is not available as a parsnip spec or when you
need a lightweight wrapper around base R. Each custom learner must provide
`fit` and `predict`, and `predict` must return one prediction per test row.

```{r custom-learners}
custom_learners <- list(
  glm = list(
    fit = function(x, y, task, weights, ...) {
      df_fit <- data.frame(y = y, x, check.names = FALSE)
      stats::glm(y ~ ., data = df_fit,
                 family = stats::binomial(), weights = weights)
    },
    predict = function(object, newdata, task, ...) {
      as.numeric(stats::predict(object, newdata = as.data.frame(newdata),
                                type = "response"))
    }
  )
)

cat("Custom learner names:\n")
names(custom_learners)
cat("Custom learner components (fit/predict):\n")
lapply(custom_learners, names)
```

This output confirms that each custom learner provides both a `fit` and
`predict` function. Use it with `fit_resample()` like this:
`fit_resample(..., learner = "glm", custom_learners = custom_learners)`.

### Visual diagnostics

`bioLeak` includes plotting helpers for quick checks:
- `plot_fold_balance()` shows class balance per fold
- `plot_overlap_checks()` highlights train/test overlap for a metadata column
- `plot_time_acf()` checks autocorrelation in time-series predictions

Misuse to avoid:
- plotting without predictions (fits with no successful folds)
- using `plot_overlap_checks()` when the column is not in split metadata
- using `plot_time_acf()` for non-temporal data

For classification, `plot_fold_balance()` uses the positive class recorded in
`fit@info$positive_class` or the second factor level if not set.

```{r plot-fold-balance}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_fold_balance(fit_safe)
} else {
  cat("ggplot2 not installed; skipping fold balance plot.\n")
}
```

The bar chart shows the number of positives and negatives in each fold, while
the dashed line shows the positive-class proportion. Large swings in the line
indicate poor stratification and can make fold-level metrics unstable.

```{r plot-overlap}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_overlap_checks(fit_leaky, column = "subject")
  plot_overlap_checks(fit_safe, column = "subject")
} else {
  cat("ggplot2 not installed; skipping overlap plots.\n")
}
```

The overlap plot compares the number of unique subjects in train vs test for
each fold. Any non-zero overlap means subject leakage. Use this plot to confirm
that subject IDs never appear in both sets simultaneously.

### Audit leakage with `audit_leakage()`

`audit_leakage()` combines four diagnostics in one object:
- permutation gap: tests whether model signal is non-random
- batch or study association with folds (chi-square and Cramers V)
- target leakage scan: feature-wise outcome association to flag proxy columns
- duplicate and near-duplicate detection using feature similarity

Assumptions and misuse to avoid:
- `coldata` must align with the original sample order
- `X_ref` must match the rows used in training (same order)
- `plot_perm_distribution()` requires `return_perm = TRUE`
- `target_threshold` should be set high enough to flag only strong proxies

**Interpretation Note:** The **Permutation Significance Test** determines if your model's
performance is statistically distinguishable from random guessing. A large gap means your
model found a signal. To determine if that signal is *real* or *leaked*, check the
**Batch Association** (confounding), **Target Leakage Scan** (proxy features),
and **Duplicate Detection** (memorization) tables.

Use `feature_space = "rank"` to compare samples by rank profiles, and
`sim_method` (`cosine` or `pearson`) to control similarity. For large `n`, `nn_k`
and `max_pairs` limit duplicate searches. Use `ci_method = "bootstrap"` with
`boot_B` if you want a confidence interval for the permutation gap.

**Leakage-aware audit**

```{r audit-leakage}
X_ref <- df[, predictors]
X_ref[c(1, 5), ] <- X_ref[1, ]

audit <- audit_leakage(
  fit_safe,
  metric = "auc",
  B = 200,
  perm_stratify = TRUE,
  batch_cols = c("batch", "study"),
  X_ref = X_ref,
  sim_method = "cosine",
  sim_threshold = 0.995,
  return_perm = TRUE
)

cat("Leakage audit summary:\n")
summary(audit)
if (!is.null(audit@permutation_gap) && nrow(audit@permutation_gap) > 0) {
  # Permutation significance results
  audit@permutation_gap
}
if (!is.null(audit@batch_assoc) && nrow(audit@batch_assoc) > 0) {
  # Batch/study association with folds (Cramer's V)
  audit@batch_assoc
} else {
  cat("No batch or study associations detected.\n")
}
if (!is.null(audit@target_assoc) && nrow(audit@target_assoc) > 0) {
  # Top features by target association score
  head(audit@target_assoc)
} else {
  cat("No target leakage scan results available.\n")
}
if (!is.null(audit@duplicates) && nrow(audit@duplicates) > 0) {
  # Top duplicate/near-duplicate pairs by similarity
  head(audit@duplicates)
} else {
  cat("No near-duplicates detected.\n")
}
```

The permutation table reports the observed metric, the mean under random label
permutation, the gap (difference), and a permutation p-value. Larger gaps (for
metrics where higher is better) indicate stronger non-random signal. The batch
association table reports chi-square statistics and Cramer's V (0 to 1), where
larger V values indicate stronger confounding between folds and batch/study
labels. The target scan table lists features with the strongest outcome
associations; for numeric features the score is |AUC - 0.5| * 2 (classification)
or |correlation| (regression), and for categorical features it is Cramer's V or
eta-squared. Scores closer to 1 indicate stronger outcome association and can
signal proxy leakage. The duplicate table lists pairs of samples with near-
identical profiles (similarity near 1), which can indicate memorization leakage.
Only the top rows are shown for the target scan and duplicate tables to keep the
output compact. Use the batch/target/duplicate tables to diagnose leakage, not
the permutation gap alone.

```{r plot-perm}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_perm_distribution(audit)
} else {
  cat("ggplot2 not installed; skipping permutation plot.\n")
}
```

The histogram shows the null distribution of the metric under label
permutation. The observed metric (vertical line) should lie well outside the
null distribution to indicate a statistically distinguishable signal.

**Audit per learner**

```{r audit-by-learner}
if (requireNamespace("ranger", quietly = TRUE)) {
  spec_rf <- parsnip::rand_forest(
    mode = "classification",
    trees = 500
  ) |>
    parsnip::set_engine("ranger")

  fit_multi <- fit_resample(
    df,
    outcome = "outcome",
    splits = safe_splits,
    learner = list(glm = spec, rf = spec_rf),
    metrics = "auc"
  )

  audits <- audit_leakage_by_learner(fit_multi, metric = "auc", B = 100)
  cat("Per-learner audit summary:\n")
  audits
} else {
  cat("ranger not installed; skipping per-learner audit example.\n")
}
```

This example uses `ranger` for the random forest spec; if it is not installed,
the chunk is skipped.

Use `parallel_learners = TRUE` to audit each learner concurrently when
`future.apply` is available.

The printed table summarizes each learner's observed metric, permutation gap,
and the strongest batch/duplicate signals. Use it to compare models on the same
leakage diagnostics, and report the model with both strong signal and minimal
confounding.

**HTML audit report**

`audit_report()` accepts either a `LeakAudit` or a `LeakFit`. When given a
`LeakFit`, it runs `audit_leakage()` first and forwards additional arguments.

```{r audit-report, eval = FALSE}
if (requireNamespace("rmarkdown", quietly = TRUE) && rmarkdown::pandoc_available()) {
  report_path <- audit_report(audit, output_dir = ".")
  cat("HTML report written to:\n", report_path, "\n")
} else {
  cat("rmarkdown or pandoc not available; skipping audit report rendering.\n")
}
```

The report path points to a standalone HTML file containing the same audit
tables and plots, suitable for sharing or archiving alongside results.

### Time-series leakage checks

Time-series data require special handling. Random splits can leak information
from the future into the past. Use `mode = "time_series"` with a prediction
`horizon`, and audit with block permutations.

**Leaky example: lookahead feature**

```{r time-series-fit}
time_splits <- make_splits(
  df_time,
  outcome = "outcome",
  mode = "time_series",
  time = "time",
  v = 4,
  horizon = 1
)

cat("Time-series splits summary:\n")
time_splits

fit_time_leaky <- fit_resample(
  df_time,
  outcome = "outcome",
  splits = time_splits,
  learner = spec,
  metrics = "auc"
)

cat("Time-series leaky fit summary:\n")
summary(fit_time_leaky)
```

The summary reports time-series cross-validated AUC. Because this example
includes a lookahead feature, the metrics can be inflated and should be treated
as a cautionary benchmark rather than a valid estimate.

**Leakage-safe alternative: remove lookahead and audit with blocks**

```{r time-series-safe}
df_time_safe <- df_time
df_time_safe$leak_future <- NULL

fit_time_safe <- fit_resample(
  df_time_safe,
  outcome = "outcome",
  splits = time_splits,
  learner = spec,
  metrics = "auc"
)

cat("Time-series safe fit summary:\n")
summary(fit_time_safe)

audit_time <- audit_leakage(
  fit_time_safe,
  metric = "auc",
  B = 200,
  time_block = "stationary",
  block_len = 10
)

cat("Time-series leakage audit summary:\n")
summary(audit_time)
if (!is.null(audit_time@permutation_gap) && nrow(audit_time@permutation_gap) > 0) {
  # Time-series permutation significance results
  audit_time@permutation_gap
}

if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_time_acf(fit_time_safe, lag.max = 20)
} else {
  cat("ggplot2 not installed; skipping ACF plot.\n")
}
```

The safe fit summary provides the leakage-resistant performance estimate. The
time-series audit uses block permutations to respect temporal dependence; a
large gap indicates non-random signal, not necessarily leakage.

The ACF plot shows autocorrelation of out-of-fold predictions by lag. Large,
systematic autocorrelation can signal temporal leakage or model misspecification
in time-series settings.

## Parallel Processing

`bioLeak` uses the `future` framework for parallelism. This works on Windows,
macOS, and Linux. To speed up simulations or audits, simply set a plan:

```{r parallel-setup, eval=FALSE}
library(future)

# Use multiple cores (works on all OS)
plan(multisession, workers = 4)

# Run a heavy simulation
sim <- simulate_leakage_suite(..., parallel = TRUE)

# Return to sequential processing
plan(sequential)
```

This chunk only configures the parallel plan, so it does not produce a result
object. Use it as a template before running compute-heavy functions.

### Simulation suite

`simulate_leakage_suite()` runs Monte Carlo simulations that inject specific
leakage mechanisms and then evaluates detection with the audit pipeline. It is
useful for validating your leakage checks before applying them to real data.
Available leakage types include `none`, `subject_overlap`, `batch_confounded`,
`peek_norm`, and `lookahead`. Use `signal_strength` to control difficulty.

```{r simulate-suite}
if (requireNamespace("glmnet", quietly = TRUE)) {
  sim <- simulate_leakage_suite(
    n = 200,
    p = 10,
    mode = "subject_grouped",
    learner = "glmnet",
    leakage = "subject_overlap",
    seeds = 1:3,
    B = 200,
    parallel = FALSE,
    signal_strength = 1
  )
  
  # Simulation results (first 6 rows)
  head(sim)
} else {
  cat("glmnet not installed; skipping simulation suite example.\n")
}
```

Each row corresponds to one simulation seed. `metric_obs` is the observed
performance (AUC for classification), `gap` is the permutation gap, and
`p_value` is the permutation-based significance. Use `metric_obs` for effect
size, and `gap/p_value` to assess whether the signal is distinguishable from
random in the simulated leakage scenario.

### Objects and summaries

bioLeak uses S4 classes to capture provenance and diagnostics:
- `LeakSplits`: returned by `make_splits()`, printed with `show()`
- `LeakFit`: returned by `fit_resample()`, summarized with `summary()`
- `LeakAudit`: returned by `audit_leakage()`, summarized with `summary()`
- `LeakAuditList`: returned by `audit_leakage_by_learner()`, printed directly
- `GuardFit`: returned by `.guard_fit()`, printed and summarized

These objects store hashes and metadata that help reproduce and audit the
workflow. Inspect their slots (`@metrics`, `@predictions`, `@permutation_gap`,
`@duplicates`) to drill into specific leakage signals.
